---
title: "Economic Drivers Of Real Estate Pricing Nationwide"
author: 'Group 10: Monirul Islam, Eric Korshun, Opal Lynch, Charles Ramirez, Jisoo
  Pak'
date: "2024-05-15"
output:
  html_document: default
  pdf_document: default
---

## Animating Question
Which economic event(s), or conditions, are correlated with real estate price fluctuations since the year 2000?

## Additional Questions 
- Is there a correlation between population growth and changes in home value? 
  - How does demand fluctuate for single-family homes?
- Does changes in personal income and consumer spending correlate with changes in home value?
- How have major events affected population dynamics, housing markets, and economic indicators within affected regions?

### Tidying Datasets 
Tidying Census Population Estimates data
``` {r include=FALSE}
library(tidyverse)
```
``` {r}
# Population Estimates 2000-2010
pop2000_2010 <- read.csv(("~/Desktop/Weylandt Project (Housing)/Raw Data /Population Estimate 2000_2010.csv")) |> 
                filter(SEX == 0) |>
                filter(AGE == 999) |>
                select(-SEX,-AGE,-DIVISION,-STATE,-ESTIMATESBASE2000,-CENSUS2010POP) |>
                filter(NAME != 'United States')
glimpse(pop2000_2010)


# Population Estimates 2011-2019
pop2011_2019 <- read.csv("~/Desktop/Weylandt Project (Housing)/Raw Data /Population Estimates 2010_2019.csv") |>
                select(-SUMLEV,-DIVISION,-STATE,-CENSUS2010POP,-ESTIMATESBASE2010,-POPESTIMATE2010) |>
                select(NAME,POPESTIMATE2011:POPESTIMATE2019) |>
                filter(NAME != 'United States') |>
                filter(NAME != 'Northeast Region') |>
                filter(NAME != 'Midwest Region') |>
                filter(NAME != 'West Region') |>
                filter(NAME != 'South Region') |>
                filter(NAME != 'Puerto Rico')
glimpse(pop2011_2019)


# Population Estimates 2020-2023
pop2020_2023 <- read.csv("~/Desktop/Weylandt Project (Housing)/Raw Data /Population Estimate 2020_2023.csv") |>
                select(NAME:POPESTIMATE2023) |>
                filter(NAME != 'United States') |>
                filter(NAME != 'Northeast Region') |>
                filter(NAME != 'Midwest Region') |>
                filter(NAME != 'West Region') |>
                filter(NAME != 'South Region') |>
                filter(NAME != 'Puerto Rico') |>
                filter(NAME != 'New England') |>
                filter(NAME != 'Middle Atlantic') |>
                filter(NAME != 'East North Central') |>
                filter(NAME != 'West North Central') |>
                filter(NAME != 'South Atlantic') |>
                filter(NAME != 'East South Central') |>
                filter(NAME != 'West South Central') |>
                filter(NAME != 'Mountain') |>
                filter(NAME != 'Pacific') |>
                select(-ESTIMATESBASE2020)
glimpse(pop2020_2023)


# Population Estimates of all states from 2000-2023
pop_estimate <- left_join(pop2000_2010,pop2011_2019,join_by(NAME==NAME)) |> 
                left_join(pop2020_2023,join_by(NAME==NAME)) |>
                rename('STATE'=NAME)
glimpse(pop_estimate)


# Separating Region and State to join other data sets 
region_state <- pop_estimate |>
                select(REGION,STATE)
glimpse(region_state)


# Population Estimates by Region 
allRegions_PopEstimates <- pop_estimate |>
                           select(-STATE) |>
                           group_by(REGION) |>
                           summarise_all(mean) |>
                           pivot_longer(cols = "POPESTIMATE2000":"POPESTIMATE2023",
                           names_to = c("YearlyPopEstimate"),
                           values_to = "Population")
allRegions_PopEstimates$Year <- gsub("POPESTIMATE(\\d{4})", "\\1", allRegions_PopEstimates$YearlyPopEstimate)
allRegions_PopEstimates <- allRegions_PopEstimates |> select(-YearlyPopEstimate)
glimpse(allRegions_PopEstimates)


# Convert To CSV
#write.csv(allRegions_PopEstimates, 'allRegions_PopEstimates.csv', row.names=FALSE)
```

Tidying Zillow Average Home Prices data
``` {r include=FALSE}
library(tidyverse)
```
``` {r}
# Original Zillow data  
Zillow <- read.csv("~/Desktop/Weylandt Project (Housing)/Raw Data /Zillow.csv") |>
          select(-RegionID,-SizeRank,-RegionType,-StateName) 
Zillow[is.na(Zillow)] <- 0
glimpse(Zillow)


# Averaging columns horizontally to consolidate values by year 
years <- 2000:2023
for (year in years) {
  monthly_cols <- grep(paste0("X", year), names(Zillow), value = TRUE)
  Zillow[[paste0(year, "_Yearly_Avg")]] <- rowMeans(Zillow[monthly_cols], na.rm = TRUE)
}


# Zillow data formatted 
Zillow_new <- Zillow |>
              select(RegionName,"2000_Yearly_Avg":"2023_Yearly_Avg") |>
              rename('STATE'=RegionName) |>
              inner_join(region_state, join_by("STATE"=="STATE")) |> # From Census Population Estimate data 
              select(REGION,STATE,"2000_Yearly_Avg":"2023_Yearly_Avg")
glimpse(Zillow_new)


# Zillow Average Home Prices by Region 
allRegions_Zillow <- Zillow_new |>
                     select(-STATE) |>
                     group_by(REGION) |>
                     summarise_all(mean) |>
                     pivot_longer(cols = "2000_Yearly_Avg":"2023_Yearly_Avg",
                     names_to = c("Year_Column"),
                     values_to = "ZillowAvgHomePrice")
allRegions_Zillow$Year <- gsub("(\\d{4})_Yearly_Avg", "\\1", allRegions_Zillow$Year_Column)
allRegions_Zillow <- allRegions_Zillow |> select(-Year_Column)
glimpse(allRegions_Zillow)


# Convert to CSV
#write.csv(allRegions_Zillow, 'allRegions_Zillow.csv', row.names=FALSE) 
```

Tidying Regional Data from The U.S. Bureau of Economic Analysis (BEA)
``` {r include=FALSE}
library(tidyverse) 
```
``` {r}
# Imported BEA Data
BEA_unclean <- read_csv('~/Desktop/Weylandt Project (Housing)/Raw Data /RegionalData.csv', show_col_types = FALSE)
glimpse(BEA_unclean)


# Disposable Income data extracted from BEA 
Disposable_Income <- BEA_unclean |> 
                     filter(LineCode == 6) |> 
                     select(-c('Description', 'LineCode', 'GeoFips')) |> 
                     mutate(across('2000':'2022', as.numeric)) |>
                     pivot_longer(c('2000':'2022'),
                        names_to = "Year", 
                        values_to = "Average Disposable Income", 
                        values_drop_na = FALSE) |>
                     pivot_wider(names_from = "Year", values_from = "Average Disposable Income")
colnames(Disposable_Income)[colnames(Disposable_Income) == "GeoName"] <- "STATE"
glimpse(Disposable_Income)


# Consumer Spending data extracted from BEA
Consumer_Spending <- BEA_unclean |> 
                     filter(LineCode == 7) |> 
                     select(-c('Description', 'LineCode', 'GeoFips')) |> 
                     mutate(across('2000':'2022', as.numeric)) |>
                     pivot_longer(c('2000':'2022'),
                        names_to = "Year", 
                        values_to = "Personal Consumption Expenditures", 
                        values_drop_na = FALSE) |>
                     pivot_wider(names_from = "Year", values_from = "Personal Consumption Expenditures")
colnames(Consumer_Spending)[colnames(Consumer_Spending) == "GeoName"] <- "STATE"
glimpse(Consumer_Spending)


# Disposable Income data by Region
allRegions_DisposableIncome <- Disposable_Income |>
                               inner_join(region_state, join_by("STATE"=="STATE")) |> # From Census Population Estimate data 
                               select(REGION,'2000':'2022') |>
                               group_by(REGION) |>
                               summarise_all(mean) |>
                               pivot_longer(cols = "2000":"2022",
                               names_to = c("Year_Column"),
                               values_to = "Disposable_Income")
allRegions_DisposableIncome$Year <- gsub("POPESTIMATE(\\d{4})", "\\1", allRegions_DisposableIncome$Year_Column)
allRegions_DisposableIncome <- allRegions_DisposableIncome |> select(-Year_Column)
glimpse(allRegions_DisposableIncome)


# Consumer Spending data by Region
allRegions_ConsumerSpending <- Consumer_Spending |>
                               inner_join(region_state, join_by("STATE"=="STATE")) |> # From Census Population Estimate data 
                               select(REGION,'2000':'2022') |>
                               group_by(REGION) |>
                               summarise_all(mean) |>
                               pivot_longer(cols = "2000":"2022",
                               names_to = c("Year_Column"),
                               values_to = "Consumer_Spending")
allRegions_ConsumerSpending$Year <- gsub("POPESTIMATE(\\d{4})", "\\1", allRegions_ConsumerSpending$Year_Column)
allRegions_ConsumerSpending <- allRegions_ConsumerSpending |> select(-Year_Column)
glimpse(allRegions_ConsumerSpending)


# Convert to CSV
#write.csv(allRegions_DisposableIncome, 'allRegions_DisposableIncome.csv', row.names=FALSE)
#write.csv(allRegions_ConsumerSpending, 'allRegions_ConsumerSpending.csv', row.names=FALSE)
```

Tidying Sales of Single-Family Homes data 
``` {r include=FALSE}
library(tidyverse)
```
``` {r}
# Sales of Single-Family Homes - Northeast Region 
Northeast_SalesSF <- read_csv('~/Desktop/Weylandt Project (Housing)/Raw Data /New One Family Houses Sold in Northeast Census Region, Thousands, Seasonally Adjusted Annual Rate .csv', show_col_types = FALSE) |>
                     rename('NorthEast'= HSN1FNE) |>
                     rename('Year'= DATE) 
Northeast_SalesSF$Year <- format(Northeast_SalesSF$Year, "%Y")
Northeast_SalesSF$NorthEast <- round(Northeast_SalesSF$NorthEast * 1000)
Northeast_Sales_new <- Northeast_SalesSF |>
                       pivot_wider(names_from = Year, values_from = NorthEast) |>
                       cbind(REGION = '1') |>
                       select(REGION, '2000':'2023')
glimpse(Northeast_Sales_new)


# Sales of Single-Family Homes - Midwest Region 
Midwest_SalesSF <- read_csv('~/Desktop/Weylandt Project (Housing)/Raw Data /New One Family Houses Sold in Midwest Census Region, Thousands, Seasonally Adjusted Annual Rate.csv', show_col_types = FALSE) |>
                   rename('MidWest'= HSN1FMW) |>
                   rename('Year'= DATE) 
Midwest_SalesSF$Year <- format(Midwest_SalesSF$Year, "%Y")
Midwest_SalesSF$MidWest <- round(Midwest_SalesSF$MidWest * 1000)
Midwest_Sales_new <- Midwest_SalesSF |> 
                     pivot_wider(names_from = Year, values_from = MidWest) |>
                     cbind(REGION = '2') |>
                     select(REGION, '2000':'2023')
glimpse(Midwest_Sales_new)


# Sales of Single-Family Homes - South Region 
South_SalesSF <- read_csv("~/Desktop/Weylandt Project (Housing)/Raw Data /New One Family Houses Sold in South Census Region, Thousands, Seasonally Adjusted Annual Rate .csv", show_col_types = FALSE) |>
                 rename('South'= HSN1FS) |>
                 rename('Year'= DATE) 
South_SalesSF$Year <- format(South_SalesSF$Year, "%Y")
South_SalesSF$South <- round(South_SalesSF$South * 1000)
South_Sales_new <- South_SalesSF |>
                   pivot_wider(names_from = Year, values_from = South) |>
                   cbind(REGION = '3') |>
                   select(REGION, '2000':'2023')
glimpse(South_Sales_new)


# Sales of Single-Family Homes - West Region 
West_SalesSF <- read_csv('~/Desktop/Weylandt Project (Housing)/Raw Data /New One Family Houses Sold in West Census Region, Thousands, Seasonally Adjusted Annual Rate .csv', show_col_types = FALSE) |>
                rename('West'= HSN1FW) |>
                rename('Year'= DATE) 
West_SalesSF$Year <- format(West_SalesSF$Year, "%Y")
West_SalesSF$West <- round(West_SalesSF$West * 1000)
West_Sales_new <- West_SalesSF |>
                  pivot_wider(names_from = Year, values_from = West) |>
                  cbind(REGION = '4') |>
                  select(REGION, '2000':'2023')
glimpse(West_Sales_new)


# Sales of Single-Family Homes - All Regions 
allRegions_SalesSFH <- bind_rows(Northeast_Sales_new,Midwest_Sales_new,South_Sales_new,West_Sales_new) |>
                       pivot_longer(cols = "2000":"2023",
                       names_to = c("Year_Column"),
                       values_to = "SFH_Sales")
allRegions_SalesSFH$Year <- gsub("(\\d{4})_Yearly_Avg", "\\1", allRegions_SalesSFH$Year_Column)
allRegions_SalesSFH <- allRegions_SalesSFH |> select(-Year_Column)
glimpse(allRegions_SalesSFH)


# Convert to CSV
#write.csv(allRegions_SalesSFH, 'allRegions_SalesSFH.csv', row.names=FALSE)
```

### Question 1 
### Is there a correlation between population growth and changes in home value? 
``` {r include=FALSE}
library(tidyverse)
```
``` {r}
allRegions_PopEstimates <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_PopEstimates.csv", show_col_types = FALSE)
allRegions_Zillow <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_Zillow.csv", show_col_types = FALSE)


# Joining All Regions Population Estimates data and All Regions Zillow Average Home Prices data 
Zillow_PopulationEstimates <- inner_join(allRegions_Zillow,allRegions_PopEstimates, join_by(REGION,Year))
glimpse(Zillow_PopulationEstimates)


# Visualization of "Changes In Home Prices and Population Estimates from 2000-2023" 
ggplot(Zillow_PopulationEstimates, aes(x=Population, y=ZillowAvgHomePrice, color=Year)) +
  geom_point() +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::comma) +
  facet_grid(~REGION, labeller = labeller(REGION = c("1" = "Northeast", "2" = "Midwest", "3" = "South", "4" = "West"))) +
  theme_bw() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Zillow Average Home Price") +
  xlab("Population Estimates") +
  ggtitle("Changes In Home Price And Population Estimates (2000-2023)") +
  scale_color_gradient(low="blue", high="red") +
  theme(legend.key.width = unit(1.5,'cm'))


# Correlation test between Population Estimates and Zillow Average Home Prices
cor.test(Zillow_PopulationEstimates$ZillowAvgHomePrice,Zillow_PopulationEstimates$Population)


# All States Population Estimates 
All_PopEstimates_longer <- pop_estimate |>
          pivot_longer(cols = "POPESTIMATE2000":"POPESTIMATE2023",
          names_to = c("YearlyPopEstimate"),
          values_to = "Population")
All_PopEstimates_longer$Year <- gsub("POPESTIMATE(\\d{4})", "\\1", All_PopEstimates_longer$YearlyPopEstimate)
All_PopEstimates_longer <- All_PopEstimates_longer |> select(-YearlyPopEstimate)
glimpse(All_PopEstimates_longer)
# All States Zillow Average Home Prices 
All_Zillow_Longer <- Zillow_new |>
          pivot_longer(cols = "2000_Yearly_Avg":"2023_Yearly_Avg",
                       names_to = c("Year_Column"),
                       values_to = "ZillowAvgHomePrice")
All_Zillow_Longer$Year <- gsub("(\\d{4})_Yearly_Avg", "\\1", All_Zillow_Longer$Year_Column)
All_Zillow_Longer <- All_Zillow_Longer |> select(-Year_Column)
glimpse(All_Zillow_Longer)


# Region 3 - South 
    All_South_Population <- All_PopEstimates_longer |> filter(REGION == 3)
    # South Population Visual
    ggplot(All_South_Population, aes(x=Year,y=Population)) +
      geom_point(aes(color=STATE)) +
      scale_y_continuous(labels = scales::comma) +
      scale_x_discrete(labels = NULL) +
      facet_wrap(~STATE) +
      theme(legend.position = 'none') +
      ggtitle("Changes In South Population Estimates (2000-2023)")
    # South Population Percent Change 
    All_SouthPop_PerChg_2000to2023 <- All_South_Population |> 
          filter(Year %in% c(2000,2023)) |> 
          group_by(STATE) |>
          summarise(Percent_Change = (Population[Year == 2023] / Population[Year == 2000] - 1) * 100)
    All_SouthPop_PerChg_2000to2023
    
    All_South_Zillow <- All_Zillow_Longer |> filter(REGION == 3)
    # South Home Price Visual 
    ggplot(All_South_Zillow, aes(x=Year,y=ZillowAvgHomePrice)) +
      geom_point(aes(color=STATE)) +
      scale_y_continuous(labels = scales::dollar) +
      scale_x_discrete(labels = NULL) +
      facet_wrap(~STATE) +
      theme(legend.position = 'none') +
      ylab("Zillow Average Home Price") +
      ggtitle("Changes In South Home Values (2000-2023)")
    # South Home Price Percent Change 
    All_SouthZillow_PerChg_2000to2023 <- All_South_Zillow |> 
          filter(Year %in% c(2000,2023)) |> 
          group_by(STATE) |>
          summarise(Percent_Change = (ZillowAvgHomePrice[Year == 2023] / ZillowAvgHomePrice[Year == 2000] - 1) * 100)
    All_SouthZillow_PerChg_2000to2023
    

# Region 4 - West
    All_West_Population <- All_PopEstimates_longer |> filter(REGION == 4)
    # West Population Visual
    ggplot(All_West_Population, aes(x=Year,y=Population)) +
      geom_point(aes(color=STATE)) +
      scale_y_continuous(labels = scales::comma) +
      scale_x_discrete(labels = NULL) +
      facet_wrap(~STATE) +
      theme(legend.position = 'none') +
      ggtitle("Changes In West Population Estimates (2000-2023)")
    # West Population Percent Change 
    All_WestPop_PerChg_2000to2023 <- All_West_Population |> 
          filter(Year %in% c(2000,2023)) |> 
          group_by(STATE) |>
          summarise(Percent_Change = (Population[Year == 2023] / Population[Year == 2000] - 1) * 100)
    All_WestPop_PerChg_2000to2023
    
    All_West_Zillow <- All_Zillow_Longer |> filter(REGION == 4)
    # West Home Price Visual 
    ggplot(All_West_Zillow, aes(x=Year,y=ZillowAvgHomePrice)) +
      geom_point(aes(color=STATE)) +
      scale_y_continuous(labels = scales::dollar) +
      scale_x_discrete(labels = NULL) +
      facet_wrap(~STATE) +
      theme(legend.position = 'none') +
      ylab("Zillow Average Home Price") +
      ggtitle("Changes In West Home Values (2000-2023)") 
    # West Home Price Percent Change 
    All_WestZillow_PerChg_2000to2023 <- All_West_Zillow |> 
          filter(Year %in% c(2000,2023)) |> 
          group_by(STATE) |>
          summarise(Percent_Change = (ZillowAvgHomePrice[Year == 2023] / ZillowAvgHomePrice[Year == 2000] - 1) * 100)
    All_WestZillow_PerChg_2000to2023
```

### How does demand fluctuate for single-family homes?
``` {r include=FALSE}
library(tidyverse)
```
``` {r}
allRegions_SalesSFH <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_SalesSFH.csv", show_col_types = FALSE)


# Visualization of Sales Of Single-Family Homes 
ggplot(allRegions_SalesSFH, aes(x=Year,y=SFH_Sales,color=REGION)) +
  geom_line() +
  scale_y_continuous(labels = scales::comma) +
  facet_grid(~REGION, labeller = labeller(REGION = c("1" = "Northeast", "2" = "Midwest", "3" = "South", "4" = "West"))) +
  ylab("Single-Family House Sales") +
  xlab("Years") +
  ggtitle("Changes In Single-Family House Sales (2000-2023)") +
  theme_gray() +
  theme(legend.position = 'none') 
```

### Question 2
### Does changes in personal income and consumer spending correlate with changes in home value?
``` {r include=FALSE}
library(tidyverse)
```

#### Disposable Income and Home Value 
``` {r}
allRegions_Zillow <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_Zillow.csv", show_col_types = FALSE)
allRegions_DisposableIncome <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_DisposableIncome.csv", show_col_types = FALSE)


# Joining All Regions Zillow Average Home Price data and All Regions Disposable Personal Income data
Zillow_DisposableIncome=inner_join(allRegions_Zillow,allRegions_DisposableIncome,join_by('REGION','Year'))
glimpse(Zillow_DisposableIncome)


# Visualization of Changes In Home Price And Disposable Income
ggplot(Zillow_DisposableIncome,aes(x=Disposable_Income,y=ZillowAvgHomePrice,color=Year))+
  geom_point() + 
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::dollar) +
  facet_grid(~REGION, labeller = labeller(REGION = c("1" = "Northeast", "2" = "Midwest", "3" = "South", "4" = "West"))) +
  theme_bw() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Zillow Average Home Price") +
  xlab("Disposable Income") +
  ggtitle("Changes In Home Price And Disposable Income (2000-2022)") +
  scale_color_gradient(low="blue", high="red") +
  theme(legend.key.width = unit(1.5,'cm'))


# Correlation test between Disposable Income and Zillow Average Home Prices
cor.test(Zillow_DisposableIncome$Disposable_Income,Zillow_DisposableIncome$ZillowAvgHomePrice)


# Permutation test to further test the correlation
    # Null Distribution - intended to have no relationship by design 
    replicate(5000, {
      Zillow_DisposableIncome |>
        mutate(ZillowAvgHomePrice = sample(ZillowAvgHomePrice)) |>
        summarize(corr=cor(ZillowAvgHomePrice,Disposable_Income)) |>
        pull(corr)
    }) -> T_corr
    hist(T_corr)
    
    # Observed (Same number from the correlation test above)
    Zillow_DisposableIncome |>
    #mutate(ZillowAvgHomePrice = sample(ZillowAvgHomePrice)) |>
    summarize(corr=cor(ZillowAvgHomePrice,Disposable_Income)) |>
    pull(corr)
# When the observed is very inconsistent with the null distribution, we can reject the null distribution. The observed data is so unlikely to have occurred if the null were true. 
```

#### Consumer Spending and Home Value 
``` {r}    
allRegions_Zillow <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_Zillow.csv", show_col_types = FALSE)
allRegions_ConsumerSpending <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_ConsumerSpending.csv", show_col_types = FALSE)


# Joining All Regions Zillow Average Home Price data and All Regions Consumer Spending data
Zillow_ConsumerSpending <- inner_join(allRegions_Zillow,allRegions_ConsumerSpending, join_by('REGION','Year'))
glimpse(Zillow_ConsumerSpending)


# Visualization of Changes In Home Price And Consumer Spending 
ggplot(Zillow_ConsumerSpending, aes(y=ZillowAvgHomePrice,x=Consumer_Spending,color=Year)) + 
  geom_point() + 
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::dollar) +
  facet_grid(~REGION, labeller = labeller(REGION = c("1" = "Northeast", "2" = "Midwest", "3" = "South", "4" = "West"))) +
  theme_bw() +
  theme(legend.position = 'bottom', axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Zillow Average Home Price") +
  xlab("Consumer Spending") +
  ggtitle("Changes In Home Price And Consumer Spending (2000-2022)") +
  scale_color_gradient(low="blue", high="red") +
  theme(legend.key.width = unit(1.5,'cm')) 


# Correlation test between Consumer Spending and Zillow Average Home Prices
cor.test(Zillow_ConsumerSpending$ZillowAvgHomePrice,Zillow_ConsumerSpending$Consumer_Spending)


# Permutation test to further test the correlation
    # Null Distribution - intended to have no relationship by design 
    replicate(5000, {
      Zillow_ConsumerSpending |>
        mutate(ZillowAvgHomePrice = sample(ZillowAvgHomePrice)) |>
        summarize(corr=cor(ZillowAvgHomePrice,Consumer_Spending)) |>
        pull(corr)
    }) -> T_corr
    hist(T_corr)
    
    # Observed (Same number from the correlation test above)
    Zillow_ConsumerSpending |>
    #mutate(ZillowAvgHomePrice = sample(ZillowAvgHomePrice)) |>
    summarize(corr=cor(ZillowAvgHomePrice,Consumer_Spending)) |>
    pull(corr)
# When the observed is very inconsistent with the null distribution, we can reject the null distribution. The observed data is so unlikely to have occurred if the null were true. 
```

### Question 3 
### How have major events affected population dynamics, housing markets, and economic indicators within affected regions?

#### 2008 Crisis
``` {r include=FALSE}
library(tidyverse)
library(scales)
library(cowplot)
library(ggplot2)
library(dplyr)
library(gridExtra)
```
``` {r}
consumerSpending <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_ConsumerSpending.csv", show_col_types = FALSE)
disposableIncome <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_DisposableIncome.csv", show_col_types = FALSE)
popEstimates <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_PopEstimates.csv", show_col_types = FALSE)
salesSFH <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_SalesSFH.csv", show_col_types = FALSE)
zillowData <- read_csv("~/Desktop/Weylandt Project (Housing)/Project csv files/allRegions_Zillow.csv", show_col_types = FALSE)
mortgage15 <- read_csv("~/Desktop/Weylandt Project (Housing)/Raw Data /MORTGAGE15US.csv", show_col_types = FALSE)
mortgage30 <- read_csv("~/Desktop/Weylandt Project (Housing)/Raw Data /MORTGAGE30US.csv", show_col_types = FALSE)
unemployment<- read_csv("~/Desktop/Weylandt Project (Housing)/Raw Data /UNRATE.csv", show_col_types = FALSE)


# Merging all data-sets into one comprehensive data-frame, with emphasis on years 2005 to 2013
economic_data <- reduce(list(consumerSpending, disposableIncome, popEstimates, salesSFH, zillowData), full_join, by = c("REGION", "Year"))
economic_data_crisis <- filter(economic_data, Year >= 2005 & Year <= 2013)


# Apply region mapping
region_mapping <- c('1' = 'Northeast', '2' = 'Midwest', '3' = 'South', '4' = 'West')
economic_data_crisis$REGION <- factor(economic_data_crisis$REGION, levels = names(region_mapping), labels = region_mapping)
glimpse(economic_data_crisis)


# Home Price & Population Trends
# Create separate data frames for home prices and population
home_prices <- economic_data_crisis[c("REGION", "Year", "ZillowAvgHomePrice")]
population <- economic_data_crisis[c("REGION", "Year", "Population")]


# Plot home price trends during 2008 crisis
home_price_plot <- ggplot(home_prices, aes(x = Year, y = ZillowAvgHomePrice, color = factor(REGION))) +
  geom_line() +
  labs(title = "Home Price Trends (2005-2013)", y = "Home Price", color = "Region") +
  theme_minimal() +
  scale_y_continuous(labels = dollar)  
home_price_plot

# Home price percent change 2008 crisis
home_price_percentChange <- home_prices|>
  group_by(REGION) |>
  mutate(Previous_ZillowAvgHomePrice = lag(ZillowAvgHomePrice, 1)) |>
  mutate(Percent_Change=((ZillowAvgHomePrice-Previous_ZillowAvgHomePrice)/Previous_ZillowAvgHomePrice)*100) |>
  select(-ZillowAvgHomePrice,-Previous_ZillowAvgHomePrice)
print(home_price_percentChange, n=36)


# Plot population trends during 2008 crisis
population_plot <- ggplot(population, aes(x = Year, y = Population, color = factor(REGION))) +
  geom_line() +
  labs(title = "Population Trends (2005-2013)", y = "Population", color = "Region") +
  theme_minimal() +
  scale_y_continuous(labels = comma)  
population_plot

# Population percent change 2008 crisis
population_percentChange <- population |> 
  group_by(REGION) |> 
  mutate(Previous_pop = lag(Population, 1)) |>
  mutate(Percent_Change=((Population-Previous_pop)/Previous_pop)*100) |>
  select(-Population,-Previous_pop)
print(population_percentChange, n=36)


# Arrange plots side by side
multi_panel_plot <- plot_grid(home_price_plot, population_plot, ncol = 2)
multi_panel_plot


# Economic Indicators Over Time (2005-2013)
ggplot(economic_data_crisis, aes(x = Year)) +
  geom_line(aes(y = Consumer_Spending, color = "Consumer Spending (USD)")) +
  geom_line(aes(y = Disposable_Income, color = "Disposable Income (USD)")) +
  geom_line(aes(y = ZillowAvgHomePrice, color = "Home Prices (USD)")) +
  facet_wrap(~REGION) +  
  labs(title = "Economic Indicators Over Time (2005-2013)", y = "Value") +
  scale_color_manual(values = c("Consumer Spending (USD)" = "blue", 
                                "Disposable Income (USD)" = "green", 
                                "Home Prices (USD)" = "red")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Indicators")) +
  scale_y_continuous(labels = comma)


# Mortgage rates from 2000-2010
# Extracting year from the DATE column for merging
mortgage15$Year <- format(as.Date(mortgage15$DATE), "%Y")
mortgage30$Year <- format(as.Date(mortgage30$DATE), "%Y")

# Convert Year to numeric for consistency
mortgage15$Year <- as.numeric(mortgage15$Year)
mortgage30$Year <- as.numeric(mortgage30$Year)

# Aggregating average yearly mortgage rates before merging
avg_mortgage15 <- mortgage15 %>%
  group_by(Year) %>%
  summarise(AvgMortgage15Rate = mean(MORTGAGE15US))

avg_mortgage30 <- mortgage30 %>%
  group_by(Year) %>%
  summarise(AvgMortgage30Rate = mean(MORTGAGE30US))

# Merging mortgage rate data with main economic data
economic_data_crisis <- economic_data_crisis %>%
  left_join(avg_mortgage15, by = "Year") %>%
  left_join(avg_mortgage30, by = "Year")

# Filtering mortgage data from 2000 to 2010
mortgage15_filtered <- filter(mortgage15, Year >= 2000 & Year <= 2010)
mortgage30_filtered <- filter(mortgage30, Year >= 2000 & Year <= 2010)

# Plotting the mortgage rate trends from 2000 to 2010
mortgage_rate_plot <- ggplot() +
  geom_line(data = mortgage15_filtered, aes(x = DATE, y = MORTGAGE15US, color = "15-Year Mortgage Rate"), linewidth = 1) +
  geom_line(data = mortgage30_filtered, aes(x = DATE, y = MORTGAGE30US, color = "30-Year Mortgage Rate"), linewidth = 1) +
  labs(title = "Mortgage Rates (2000-2010)",
       x = "Year",
       y = "Mortgage Rate (%)",
       color = "Mortgage Rate Type") +
  scale_color_manual(values = c("15-Year Mortgage Rate" = "steelblue", "30-Year Mortgage Rate" = "darkorange")) +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
mortgage_rate_plot


# Unemployment visualization 
ggplot(data = unemployment, aes(x = DATE, y = UNRATE)) +
  geom_line() + # This adds a line graph
  labs(title = "Unemployment Rate Over Time",
       x = "Date",
       y = "Unemployment Rate (%)") +
  theme_minimal()
```

#### COVID Crisis
``` {r include=FALSE}
library(tidyverse)
library(sf)
library(tidyverse)
library(usmap)
library(gganimate)
library(gapminder)
library(gifski)
```
``` {r}
# Focus on the years from 2019 to 2022
Covid_crisis <- filter(economic_data, Year >= 2019 & Year <= 2022)


# Apply region mapping
region_mapping <- c('1' = 'Northeast', '2' = 'Midwest', '3' = 'South', '4' = 'West')
Covid_crisis$REGION <- factor(Covid_crisis$REGION, levels = names(region_mapping), labels = region_mapping)
glimpse(Covid_crisis)


# Population trends during Covid 
population_covid <- Covid_crisis[c("REGION", "Year", "Population")]

# Plot population trends Covid
population_plot_covid <- ggplot(population_covid, aes(x = Year, y = Population, color = factor(REGION))) +
  geom_line() +
  labs(title = "Population Trends (COVID)", y = "Population", color = "Region") +
  theme_minimal() +
  scale_y_continuous(labels = comma)  
population_plot_covid

# Population Percent Change during Covid 
population_percentChange_covid <- population_covid |> 
  group_by(REGION) |> 
  mutate(Previous_pop = lag(Population, 1)) |>
  mutate(Percent_Change=((Population-Previous_pop)/Previous_pop)*100) |>
  select(-Population,-Previous_pop)
population_percentChange_covid


# US shape file on R
us_map <- us_map(regions = "states") |> select(full,geom)


# Merging All States Zillow data with US shape file 
Zillow_2019_2023 <- All_Zillow_Longer |> 
  group_by(STATE) |> 
  filter(Year %in% c(2019:2023)) |>
  inner_join(us_map, join_by('STATE'=='full'))


# Animation of Changes in Zillow Average Home Prices during Covid 
ggplot(Zillow_2019_2023, aes(fill=ZillowAvgHomePrice, geometry = geom)) +
  geom_sf() +
  transition_time(as.integer(Year)) +
  theme(
    axis.title = element_blank(),  
    axis.text = element_blank(),   
    axis.ticks = element_blank(),  
    panel.background = element_blank(),  
    panel.border = element_blank(),  
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),
    legend.position = "left")+ 
  scale_fill_continuous() +
  scale_x_log10() +
  scale_y_log10() +
  scale_fill_distiller(type="div",
                       palette="RdGy",
                       name="Zillow Price Scale",
                       direction=-1,
                       labels=scales::dollar_format())  +
  guides(color="none") +
  labs(fill = "Zillow Price Scale",
       caption = "Source: Zillow Home Value Index: zillow.com/data/") +
  ggtitle("{frame_time} Zillow Average Home Price Change")


# Population Pyramid of Single-Family Houses Sales 
ggplot(allRegions_SalesSFH, aes(x=Year, y=SFH_Sales, fill=REGION)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  facet_wrap(~REGION, scales = "free_y", labeller = labeller(REGION = c("1" = "Northeast", "2" = "Midwest", "3" = "South", "4" = "West"))) +  
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Change in Sales of Single-Family Homes: 2000 - 2023", x = "Year", y = "Number of Sales") +
  theme_bw()+
  theme(legend.position = "none") +
  scale_color_date() 


# Trend analysis over Covid years
ggplot(Covid_crisis, aes(x = Year)) +
  geom_line(aes(y = Consumer_Spending, color = "Consumer Spending (USD)")) +
  geom_line(aes(y = Disposable_Income, color = "Disposable Income (USD)")) +
  geom_line(aes(y = ZillowAvgHomePrice, color = "Home Prices (USD)")) +
  facet_wrap(~REGION) +
  labs(title = "Economic Indicators Over Time (COVID)", y = "Value") +
  scale_color_manual(values = c("Consumer Spending (USD)" = "blue", 
                                "Disposable Income (USD)" = "green", 
                                "Home Prices (USD)" = "red")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Indicators")) +
  scale_y_continuous(labels = comma)  
```

